A generalized definition of multidimensional item response theory parameters

### 2024-04-09 output/paper_draft/paper_M2PLM_parameters.Rmd
Changes grammar (minor suggestion by Grammarly)

A crucial assumption leading to those formulations is that the axes of the
latent space are orthogonal
[@ackerman_using_1994; @ackerman_multidimensional_2005; see also
@reckase_difficulty_1985, p. 404 and Equation 7].
Of course, orthogonality is a property of a geometrical space instead of a
latent trait space.
Thus, nothing prevents a researcher from interpreting an abstraction such as
a latent trait space, no matter its structure, as an orthogonal space
[@ackerman_creating_1994]
---exploratory factor analysis software, such as SPSS <!--# Reference needed? -->or
the R package *psych*, <!--# Reference needed? -->graphically represents
factor loadings on orthogonal axes even after an oblique rotation.
Nevertheless, the assimilation of correlated latent traits to non-orthogonal
axes is prevalent in the common factor literature
[see, e.g., @harman_modern_1970].<!--# Thurstone's Multiple Factor Analysis, 1947? -->
This assumption is not so easily found in the MIRT literature though,
which mostly deals with uncorrelated dimensions
[see @reckase_multidimensional_2009 for a comprehensive treatment of the topic].
This fact is probably due to the different spotlights
these two theoretical programs put the focus on;
factor analysis aims at a clear interpretation of
the latent covariance structure, and thus focuses on finding
a structure matrix as simple as possible. MIRT literature, on the other hand,
aims at modeling responses to certain stimuli and estimating person scores,
being the model interpretation secondary in relevance.<!--# Reference needed? -->
Nevertheless, the confluence of the two theoretical approaches into a common
modern measurement theory [@mcdonald_test_1999] requires
the relaxation of traditional constraints on both approaches.
It is not uncommon to find recent examples of MIRT models where the
correlation matrix of the latent dimensions is freely estimated
[see, e.g., @reckase_multidimensional_2009, pp. 224-228]. Despite this, the
correspondence between correlated latent traits and orthogonal coordinate axes
is usually implicit in the MIRT literature without further consideration<!--# Examples? -->
[see @zhang_theoretical_1999 for a notable exception].
Moreover, disregarding this geometrical interpretation may lead to
misconceptions that put the measurement and its conclusions into question.<!-- completar?? -->

A latent space where the traits are correlated is easier to understand as a
geometrical space with non-orthogonal axes,
often referred to as *general Cartesian coordinates* [@harman_modern_1970].
Such a space has the advantage that the cosines of the angles between two axes
can be interpreted as the correlation of the corresponding latent dimensions.
<!--# Reference needed? -->
<!-- Problem here: The latent space is not defined like that, according to
the results of section "Linear algebra of non-orthogonal coordinates". -->
The general case of MIRT model estimation implies estimating possibly
non-null covariances among latent dimensions, as well as non-standard scales.
Assuming orthonormality among traits is usually unrealistic, be it in the
cognitive [e.g., @reckase_multidimensional_2009] or the non-cognitive domain<!--# Reference to example? -->.
Moreover, certain applications impose the estimation of
non-identity latent covariance matrices (e.g. multigroup equating) even when<!-- # Reference needed? -->
interpreting the latent dimensions as substantive constructs is not a goal.
As we will argue, the usual definitions of the MID and MDISC parameters
do not generalize to the non-orthonormal geometry
necessary to accurately represent correlated latent spaces.

In the non-cognitive domain, the use of so-called negatively-keyed items
--items that tap the negative pole of a certain trait,
usually reverse scored-- is also widespread.<!--# Reference and example? -->
It is common to find instruments that combine
positively- and negatively-keyed items.
Moreover, a complex item may have discrimination parameters with opposite signs
for different latent dimensions.<!--# Examples? -->
Take the multidimensional version of the nominal categories model
[@thissen_nominal_2010] as an example,
and consider two response categories tapping a different construct each:
A positive change in the dimension tapped by one category will inevitably
lead to a decrease in the probability of response to the other category.
One can easily note that a monotonically decreasing probability for a certain
dimension is given by a negative discrimination parameter
(and an unchanging probability by a null discrimination parameter),<!--# Actually, this also applies to the cognitive domain, which may not be very well explained. -->
contrary to the constraints imposed by @reckase_difficulty_1985 [p. 411].
However, the multidimensional item parameters can also be defined for
items with monotonically non-increasing probabilities, as we will see.

The main objective of this paper is thus to provide a generalization of
the `r MODEL_ACRONYM` multidimensional parameters in two aspects:
On one hand, in an oblique coordinate system
that can be soundly applied to the correlated trait case; on the other, to
items that yield monotonous response probabilities, but in the broader sense of
unchanging monotony. Additionally, we aim at providing a method for
graphically representing these items in the fashion of @ackerman_creating_1994
but taking the non-orthogonality of the coordinate axes into account.
In order to pursue these objectives, we first define the generalization of
the `r MODEL_ACRONYM` model and describe its scope of application.
Then, we follow the method outlined by @reckase_difficulty_1985 and
@reckase_discriminating_1991 to obtain
the formulation of the multidimensional parameters.
We will introduce two versions of these indices,
showing that either the original or one of the new ones is obtained
as a function of how we define and interpret the latent space.
Next, we discuss the properties of the indices,
propose a graphical representation method,
and present two practical examples applying them.
We round up with a discussion about
the applicability and generalization of our results.

@@

Consider a $`r N_DIMS`$-dimensional latent space $`r LATENT_SPACE`$
in basis $`r LS_BASIS_EQ`$, being $`r LS_BASIS`$ not necessarily orthonormal,
where a respondent is represented by vector $`r TRAIT_VEC_IN_LS`$.
According to the `r MODEL_ACRONYM` model, the probability of
a positive response to item $`r ITEM_INDEX`$ is

@@

We must stop here and make a few remarks. First, as we are not constraining
ourselves to cognitive traits,
we speak of _positive_ instead of _correct response_. This may imply giving
a correct response indeed, if we are talking about a cognitive test.
But it may mean endorsing an item (or a certain response option)
in a non-cognitive domain.
Second, although not explicitly stated, $`r DISCR_VECTOR`$ is a vector of
non-negative real numbers in Reckase
[-@reckase_difficulty_1985, a decrease in the probability of a correct response
with an increase in ability makes little sense].
In our case, $`r DISCR_VECTOR`$ is a vector of real numbers; the value
of the $`r DIM_INDEX`$-th component $`r DISCR_PARAM`$ can be negative if
the item is *inverse* for (i.e., taps the negative pole of)
latent dimension $`r DIM_INDEX`$ (e.g. a negatively-keyed item).
Finally, as responses are not necessarily interpreted as correct or incorrect,
parameter $`r INTERCEPT_PARAM`$ is said to be related to the item _location_
instead of the item difficulty.

Our aim is to find in the first place
the point of maximum slope of $`r IRF_ABBR`$ in $`r LATENT_SPACE`$.
Because the slope is dependent on the direction considered,
we need to reparameterize Equation \@ref(eq:M2PL-formulation)
in polar coordinates, as in @reckase_difficulty_1985.
However, Reckase's Equation 3 implicitly assumes orthogonality.
In order to contemplate the general case,
we need to introduce a few linear algebra results before proceeding further.

@@

### Test space and latent space

Let $`r TEST_SPACE`$ be the vector space made up by
the set of $`r DISCR_VECTOR`$ vectors,
with the sum and product by a scalar operation as in $`r REAL_N_SPACE`$.
Consider also an orthonormal basis $`r TEST_SPACE_STD_BASIS`$
in $`r TEST_SPACE`$, such that $`r DISCR_STD_COORDS_ALT_EQ`$.
Since the response probability to the `r MODEL_ACRONYM`,
given by \@ref(eq:M2PL-formulation), must be invariant with respect to
the change of basis expressed in \@ref(eq:change-basis),
we must define the coordinates $`r DISCR_STD_COORDS`$ of the parameter vector
$`r DISCR_VECTOR`$ in the orthonormal basis $`r TEST_SPACE_STD_BASIS`$,
such that $`r INVARIANCE_EQ`$.
If we assume that $`r DISCR_VECTOR`$ exists in vector space $`r TEST_SPACE`$,
with coordinates $`r DISCR_COORDS`$ in basis $`r TEST_SPACE_BASIS`$,
then we must define $`r DISCR_COORDS`$ such that

\begin{equation}
  `r DISCR_STD_COORDS_EQ`,                               (\#eq:discr-vector-std)
\end{equation}

We want @reckase_difficulty_1985's definition of the multidimensional parameters
in rectangular coordinates to be a particular case of our more general
definition, so we can conveniently assume that the inner product in
$`r TEST_SPACE_STD_BASIS`$ is the dot product,
as we did with $`r LS_ORTH_BASIS`$.
The norm of $`r DISCR_VECTOR`$ thus fulfills

\begin{equation}
  `r DISCR_VECTOR_MODULE_EQ`.
\end{equation}

Therefore, an inner product $`r TEST_SPACE_INNER_PROD`$ with Gram matrix
$`r INNER_PROD_MATRIX_INV`$ can be defined, such that $`r INNER_PROD_DISCR_EQ`$
keeps the norm of $`r DISCR_VECTOR`$ invariant.

We shall refer to this new space as the *test space* [@zhang_theoretical_1999].
Although the inverse relationship between their respective inner product
Gram matrices induces a bijective relationship between the two sets of spaces,
it is worth highlighting that test space $`r TEST_SPACE`$
and latent space $`r LATENT_SPACE`$ are different entities. Only in the
orthonormal case the two spaces share a common inner product
(i.e., the dot product), and thus can be mistaken.

### Direction cosines in the latent space

@@

Applying Proposition 1 to the model formulation in \@ref(eq:M2PL-formulation),
the `r MODEL_ACRONYM` model is expressed in polar coordinates as

@@

As Equation \@ref(eq:M2PL-formula-polar) shows, the inner product matrix is
involved in the expression of the model in polar coordinates.
Importantly, this shows that @reckase_difficulty_1985's Equation 3 (p. 403)
already assumes the orthonormality of the latent space, as we stated before.

@@

The slope in direction $`r DIR_ANGLE_VEC_ORG`$ is given by

@@

\begin{equation}
  \begin{split}
    `r DIR_COS_AS_DIR_COS_ORTH_EQ`, \\
    `r TRAIT_POLAR_COEFF_COS_ORTH_EQ`.
  \end{split}                                 (\#eq:direction-cosine-vectors-eq)
\end{equation}

@@

As $`r DISCR_VECTOR`$ is defined in test space $`r TEST_SPACE`$ with basis
$`r TEST_SPACE_BASIS`$, its algebraic vector values correspond to its
coordinates, i.e., $`r DISCR_VECTOR_COORDS_EQ`$. Therefore, by Equation
\@ref(eq:discr-vector-std),

@@

Substituting $`r DIR_COS_ITEM_VEC_STD`$
(Equation \@ref(eq:direction-cosine-vectors-eq)) and
$`r DISCR_STD_COORDS`$ (Equation \@ref(eq:discr-vector-std))
by their respective expressions results in the
expression for the direction cosine vector of item $i$ in $`r LATENT_SPACE`$,

@@

We have that

@@

Applying Proposition 1, in the test space $`r TEST_SPACE`$, we get

@@

The (signed) distance to the origin,
given by the first component of $`r MIL_PARAM`$, has an interesting property:
By the definition of the inner product,
$`r MDISC_SYM`$ is strictly positive for non-null vectors.
Therefore, the sign of $`r DISTANCE_PARAM`$ is equal to
the sign of $`r INTERCEPT_PARAM`$.
The implication is that the item is displaced along
its measurement direction forwards or backwards with respect to the origin.

>>>>

A crucial assumption leading to those formulations is that the axes of the
latent space are orthogonal
[@ackerman_using_1994; @ackerman_multidimensional_2005; see also
@reckase_difficulty_1985, p. 404 and Equation 7].
Of course, orthogonality is a property of a geometrical space instead of a
latent trait space.
Thus, nothing prevents a researcher from interpreting an abstraction such as
a latent trait space, no matter its structure, as an orthogonal space
---exploratory factor analysis software, such as SPSS [@ibm_corp_ibm_2022] or
the R package *psych* [@revelle_psych_2024], graphically represents
factor loadings on orthogonal axes even after an oblique rotation.
Nevertheless, the assimilation of correlated latent traits to non-orthogonal
axes is prevalent in the common factor literature
[see, e.g., @harman_modern_1970].
This assumption is not so easily found in the MIRT literature though,
which mostly deals with uncorrelated dimensions
[see @reckase_multidimensional_2009 for a comprehensive treatment of the topic].
This fact is probably due to the different spotlights
these two theoretical programs put the focus on;
factor analysis aims at a clear interpretation of
the latent covariance structure, and thus focuses on finding
a structure matrix as simple as possible. MIRT literature, on the other hand,
aims at modeling responses to certain stimuli and estimating person scores,
being the model interpretation secondary in relevance.<!--# Reference needed? -->
Nevertheless, the confluence of the two theoretical approaches into a common
modern measurement theory [@mcdonald_test_1999] requires
the relaxation of traditional constraints on both approaches.
It is not uncommon to find recent examples of MIRT models where the
correlation matrix of the latent dimensions is freely estimated
[see, e.g., @reckase_multidimensional_2009, pp. 224-228]. Despite this, the
correspondence between correlated latent traits and orthogonal coordinate axes
is usually implicit in the MIRT literature without further consideration
[see @zhang_theoretical_1999 for a notable exception].
Moreover, disregarding this geometrical interpretation may lead to
misconceptions that put the measurement and its conclusions into question.<!-- completar?? -->

A latent space where the traits are correlated is easier to understand as a
geometrical space with non-orthogonal axes,
often referred to as *general Cartesian coordinates* [@harman_modern_1970].
Such a space has the advantage that the cosines of the angles between two axes
can be interpreted as the correlation of the corresponding latent dimensions.
<!--# Reference needed? -->
<!-- Problem here: The latent space is not defined like that, according to
the results of section "Linear algebra of non-orthogonal coordinates". -->
The general case of MIRT model estimation implies estimating possibly
non-standard latent dimensions and non-null covariances among them
(a more realistic case than an identity matrix).
Assuming orthonormality among traits is usually unrealistic, be it in the
cognitive [e.g., @reckase_multidimensional_2009] or the non-cognitive domain<!--# Reference to example? -->.
Moreover, certain applications impose the estimation of
non-identity latent covariance matrices (e.g. multigroup equating) even when<!-- # Reference needed? -->
interpreting the latent dimensions as substantive constructs is not a goal.
As we will argue, the usual definitions of the multidimensional difficulty and
discrimination parameters do not generalize to the non-orthonormal geometry
necessary to accurately represent correlated latent spaces.

In the non-cognitive domain, the use of so-called negatively-keyed items
---items that tap the negative pole of a certain trait, and thus
usually reverse-scored--- is also widespread.<!--# Reference and example? -->
It is common to find instruments that combine
positively- and negatively-keyed items.
Moreover, a complex item may have discrimination parameters with opposite signs
for different latent dimensions.<!--# Examples? -->
Take the multidimensional version of the nominal categories model
[@thissen_nominal_2010] as an example,
and consider two response categories tapping a different construct each:
A positive change in the dimension tapped by one category will inevitably
lead to a decrease in the probability of response to the other category.
One can easily note that a monotonically decreasing probability for a certain
dimension is given by a non-positive discrimination parameter
(and an unchanging probability by a null discrimination parameter),<!--# Actually, this also applies to the cognitive domain, which may not be very well explained. -->
contrary to the constraints imposed by @reckase_difficulty_1985 [p. 411].
However, the multidimensional item parameters can also be defined for
items with monotonically non-increasing probabilities, as we will see.

The main objective of this paper is thus to provide a generalization of
the MIRT parameters in two aspects: on one hand, in an oblique coordinate system
that can be soundly applied to the correlated trait case; on the other, to
items that yield monotonous response probabilities, but in the broader sense of
unchanging monotony. Additionally, we aim to provide a method for
graphically representing these items in the fashion of @ackerman_creating_1994
but taking the non-orthogonality of the coordinate axes into account.
In order to pursue these objectives, we first define the generalization of
the `r MODEL_ACRONYM` model and describe its scope of application.
Then, we follow the method outlined by @reckase_difficulty_1985 and
@reckase_discriminating_1991 to obtain
the formulation of the multidimensional parameters.
We will introduce two versions of these indices,
showing that either the original or one of the new ones is obtained
as a function of how we define and interpret the latent space.
Next, we discuss the properties of the indices,
propose a graphical representation method,
and present two examples of application of our results.
We round up with a discussion about
the applicability and generalization of our results.

@@

Consider a $`r N_DIMS`$-dimensional latent space $`r LATENT_SPACE`$
in basis $`r LS_BASIS_EQ`$.
A respondent is represented by vector $`r TRAIT_VEC_IN_LS`$,
which is assumed to be a multivariate normal random variable:

\begin{equation}
  `r TRAIT_MV_DEF`,
\end{equation}

where $`r MEAN_VECTOR`$ is an $n$-element vector of means and
$`r COV_MATRIX` = `r SD_MATRIX` `r CORR_MATRIX` `r SD_MATRIX`$
an $n$-dimensional positive definite covariance matrix,
with $`r CORR_MATRIX`$ the corresponding correlation matrix
(also positive definite) and $`r SD_MATRIX`$ a scaling diagonal matrix
with $`r VAR_COV_EQ`$ (i.e., the variances).

According to the `r MODEL_ACRONYM` model, the probability of
a positive response to item $`r ITEM_INDEX`$ is

@@

We must make a few remarks here. First, as we are not constraining ourselves to
cognitive traits, we speak of _positive_ instead of _correct response_.
This may imply a correct response, if we are talking about a cognitive test, or
endorsing an item (or a certain response option) in a non-cognitive domain.
Second, although not explicitly stated, $`r DISCR_VECTOR`$ is a vector of
non-negative real numbers in Reckase
[-@reckase_difficulty_1985, a decrease in the probability of a correct response
with an increase in ability makes little sense].
In our case, $`r DISCR_VECTOR`$ is a vector of real numbers; the value
of the $`r DIM_INDEX`$-th component $`r DISCR_PARAM`$ can be negative if
the item is *inverse* for (i.e., taps the negative pole of)
latent dimension $`r DIM_INDEX`$.
Finally, as responses are not necessarily interpreted as correct or incorrect,
parameter $`r INTERCEPT_PARAM`$ is said to be related to the item _location_,
instead of the item difficulty.

Our aim is to find the point of maximum slope of $`r IRF_ABBR`$ in
$`r LATENT_SPACE`$.
Because the slope is dependent on the direction considered,
we need to reparameterize Equation \@ref(eq:M2PL-formulation)
in polar coordinates, as in @reckase_difficulty_1985.

However, Reckase's Equation 3 implicitly assumes orthogonality.
In order to contemplate the general case,
we need to introduce a few linear algebra results before proceeding further.

@@

@@

From Proposition 1,
any trait vector $`r TRAIT_VECTOR`$ is expressed in polar coordinates as

\begin{equation}
  `r TRAIT_VECTOR_POLAR_EQ`.                      (\#eq:trait-equivalence-polar)
\end{equation}

Thus, the `r MODEL_ACRONYM` model is expressed in polar coordinates as

@@

As Equation \@ref(eq:M2PL-formula-polar) shows, the inner product matrix is
involved in the expression of the model in polar coordinates.
Importantly, this shows that @reckase_difficulty_1985's Equation 3 (p. 403)
already assumes the orthonormality of the latent space.

@@

The slope in direction $`r DIR_ANGLE_VEC`$ is given by

@@

\begin{equation}
  \begin{split}
    `r DIR_COS_AS_DIR_COS_ORTH_EQ` \\
    `r TRAIT_POLAR_COEFF_COS_ORTH_EQ`.
  \end{split}                                 (\#eq:direction-cosine-vectors-eq)
\end{equation}

@@

We can then define

\begin{equation}
  `r DISCR_VECTOR_STD_EQ`                                (\#eq:discr-vector-std)
\end{equation}

so

@@

Substituting $`r DIR_COS_ITEM_VEC_STD`$
(Equation \@ref(eq:direction-cosine-vectors-eq)) and
$`r DISCR_VECTOR_STD`$ (Equation \@ref(eq:discr-vector-std))
by their respective expressions results in the
expression for the direction cosine vector of item $i$ in $`r LATENT_SPACE`$,

@@

## Test space and latent space

Before defining the multidimensional parameters, we need to make a clarification
regarding the representation of the items in the latent space.
We call $`r TEST_SPACE`$ the *test space* [@zhang_theoretical_1999] of
the discrimination vectors.
This space is where the item vectors can be properly represented.
It is in general different from the latent space $`r LATENT_SPACE`$,
although there is a bijective relationship between the two sets of spaces.
Let's consider the model given by Equation \@ref(eq:M2PL-formulation),
where the latent traits and the parameters $`r DISCR_VECTOR`$ are expressed
in the same basis $`r LS_BASIS`$ in their respective spaces.

Since the response probability must be invariant, because of Equation
\@ref(eq:change-basis), parameter $`r DISCR_VECTOR`$ must also be transformed
in $`r DISCR_VECTOR_STD`$.
Equating the probabilities we get $a_1^{'}=P^{-1^T}a_i=M^{-1}Pa_i$ .
This shows that first a basis change to the standard base is performed and then
a transformation through $`r INNER_PROD_MATRIX_INV`$,
so the $`r DISCR_VECTOR_STD`$ are expressed in the standard basis.

\begin{equation}
  `r DISCR_VECTOR_MODULE_EQ`                     (\#eq:discr-param-squared-norm)
\end{equation}

Therefore, in the test space, an inner product matrix
$`r INNER_PROD_MATRIX_INV`$ can be defined such that $`r INNER_PROD_DISCR_EQ`$,
to keep the norm invariant.
With this definition we have that

@@

Applying Proposition 1,

@@

The (signed) distance to the origin,
given by the second element of $`r MIL_PARAM`$, has an interesting property:
By the definition of the inner product,
$`r MDISC_SYM`$ is strictly positive for non-null vectors.
Therefore, the sign of $`r DISTANCE_PARAM`$ is equal to
the sign of $`r INTERCEPT_PARAM`$.
The implication is that the item is displaced along
its measurement direction forwards or backwards with respect to the origin.
